Trajectories

### 2022-11-17 index.qmd
Update lab notes

```{r}
#| label: setup
#| include: false
options(
    knitr.kable.NA = "-",
    knitr.duplicate.label = "allow",
    ggplot2.discrete.fill = ggsci::pal_d3()(4),
    ggplot2.discrete.colour = ggsci::pal_d3()(4),
    ggplot2.continuous.fill = ggplot2::scale_color_gradient,
    ggplot2.continuous.colour = ggplot2::scale_color_gradient
)
```

```{r}
#| label: prepare
#| echo: true
#| message: false
#| warning: false
# set params and load target contents

library(bayesplot)
library(brms) 
library(bayestestR)
library(conflicted)
library(dplyr) 
library(forcats)
library(ggplot2) 
library(ggsci)
library(gt) 
library(here) 
library(janitor)
library(knitr)
library(lubridate)
library(multilex)
library(patchwork) 
library(papaja) 
library(purrr)
library(scales) 
library(stringdist)
library(stringr)
library(targets)
library(tibble)
library(tidybayes) 
library(tidyr)

# load R objects
tar_load_everything()

# set custom ggplot2 theme
theme_set(theme_custom())

# resolve namespace conflicts
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
```

# Questionnaires

```{r}
#| label: questionnaire
#| echo: true
#| message: false
#| warning: false
pool <- multilex::pool %>% 
    drop_na(cognate)
n_categories <- distinct(pool, category) %>% nrow()
n_items <- length(unique(pool$item))
n_item_language <- count(pool, language) %>% pull(n) %>% set_names(c("catalan", "spanish"))
n_item_cognate <- count(pool, language, cognate) %>% 
    pull(n) %>% 
    set_names(c("catalan-cognate", "catalan-noncognate", "spanish-cognate", "spanish-noncognate"))
```

The questionnaire was implemented on-line using the formR platform [@arslan2020formr], and was structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items (`r n_item_cognate["catalan-cognate"]` cognates, `r n_item_cognate["catalan-noncognate"]` non-cognates) and the Spanish inventory contained `r n_item_language["spanish"]` (`r n_item_cognate["spanish-cognate"]` cognates, `r n_item_cognate["spanish-noncognate"]` non-cognates). Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* \[dog\] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. When there were two acceptable translation equivalents for a given word, we included both in separate items (e.g., Catalan *acabar* \[*to finish*\] and Spanish *acabar* and *terminar*), or merged them into a single items (e.g., Spanish *mono* \[*monkey*\] and Catalan *mono/mico*. We included items from a diverse sample of `r n_categories` semantic/functional categories (see Appendix 1). For the analyses included in this study, we excluded items from the adverbs, auxiliary words, connectives, interjections and games and routines categories, so that only data from content words (nouns, adjectives, and verbs) were used.

@@

# Items

```{r}
#| label: tbl-items-summary
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Lexical frequencies. Mean, standard error, and 95% confidence interval of lexical frequencies of items included in the Catalan and Spanish lists, reported separately for identical cognates, non-identical cognates, and non-cognates."
items <- left_join(items, select(multilex_data$pool, te, item, language, class, category))

items %>%
    unnest(list) %>% 
    group_by(language, list) %>% 
    summarise_at(
        vars(freq, n_phon, lv), 
        lst(n = ~sum(!is.na(.)), mean, sd, min, max)
    ) %>% 
    select(-c(n_phon_n, lv_n)) %>% 
    rename(n = freq_n) %>% 
    gt(groupname_col = "language") %>% 
    fmt_number(matches("freq_|n_phon_mean|n_phon_sd")) %>% 
    fmt_percent(matches("lv_")) %>% 
    tab_spanner("Frequency (Zipf)", columns = starts_with("freq_")) %>% 
    tab_spanner("# Phonemes", columns = starts_with("n_phon_")) %>% 
    tab_spanner("Levenshtein", columns = starts_with("lv_")) %>% 
    cols_merge_range(col_begin = "freq_min", col_end = "freq_max") %>% 
    cols_merge_range(col_begin = "n_phon_min", col_end = "n_phon_max") %>% 
    cols_merge_range(col_begin = "lv_min", col_end = "lv_max") %>% 
    cols_label(
        list = "",
        n = md("*N*"),
        freq_mean = md("Mean"),
        freq_sd = md("*SD*"),
        freq_min = md("Range"),
        n_phon_mean = md("Mean"),
        n_phon_sd = md("*SD*"),
        n_phon_min = md("Range"),
        lv_mean = md("Mean"),
        lv_sd = md("*SD*"),
        lv_min = md("Range")
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(everything())
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(everything())
    )
```

## Lexical frequency

```{r}
#| label: fig-items-freq
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Distribution of lexical frequency across items"
items %>% 
    ggplot() +
    aes(
        x = freq,
        fill = language,
        colour = language,
        shape = language
    ) +
    stat_slab(
        slab_colour = "white",
        slab_size = 1,
        alpha = 0.75,
    ) +
    stat_pointinterval(
        aes(y = -0.1),
        position = position_dodge(width = 0.1),
        point_size = 3
    ) +
    geom_hline(yintercept = 0, colour = "black", size = 0.5) +
    labs(
        x = "Frequency (Zipf)",
        y = "Probability density",
        fill = "Language",
        shape = "Language",
        colour = "Language"
    ) +
    scale_y_continuous(labels = percent,  limits = c(-0.15, 1), breaks = seq(0, 1, 0.25)) +
    theme(
        legend.position = "top",
        legend.title = element_blank()
    )
```

## \# Phonemes

```{r}
#| label: fig-items-nphon
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Distribution of number of phonemes across items"
items %>%
    count(language, n_phon) %>% 
    mutate(prop = n/sum(.$n)) %>% 
    ggplot() +
    aes(
        x = n_phon,
        y = prop,
        fill = language,
        colour = language,
        shape = language
    ) +
    geom_col(
        position = position_dodge(width = 0.95)
    ) +
    geom_hline(yintercept = 0, colour = "black", size = 0.5) +
    labs(
        x = "# Phonemes",
        y = "Probability density",
        fill = "Language",
        shape = "Language",
        colour = "Language"
    ) +
    scale_y_continuous(labels = percent, breaks = seq(0, 1, 0.025)) +
    scale_x_continuous(breaks = seq(0, 20, 1)) +
    theme(
        legend.position = "top",
        axis.ticks.x = element_line(),
        legend.title = element_blank()
    )
```

## Levenshtein

```{r}
#| label: fig-items-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Distribution of Levenshtein similarities across items"
items %>% 
    distinct(te, .keep_all = TRUE) %>% 
    mutate(lv_cut = cut(lv, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>% 
    count(lv_cut) %>% 
    mutate(prop = n/sum(.$n)) %>% 
    ggplot() +
    aes(x = lv_cut, y = prop) +
    geom_col(fill = ggsci::pal_d3()(4)[1]) +
    scale_y_continuous(labels = percent, breaks = seq(0, 1, 0.025)) +
    labs(
        x = "Levenshtein",
        y = "Probability density",
    ) +
    theme(
        legend.position = "top"
    )
```

## Frequency by \# Phonemes

```{r}
#| label: fig-items-frequency-phonemes
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association betwen number of lexical frequency and number of phonemes"
items %>% 
    drop_na(freq) %>% 
    ggplot(aes(n_phon, freq, colour = language, fill = language)) +
    facet_wrap(~language) +
    geom_point(size = 1, alpha = 0.25) +
    geom_smooth(method = "lm", formula = "y ~ x") +
    labs(
        x = "# Phonemes", 
        y = "Lexical frequency (Zipf score)\nExtracted from SUBTLEX", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    ) 
```

## Frequency by Levenshtein

```{r}
#| label: fig-items-frequency-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association betwen lexical frequency and phonological similarity"
items %>% 
    left_join(select(pool, item, language)) %>% 
    drop_na(freq) %>% 
    ggplot(aes(freq, lv, colour = language, fill = language)) +
    facet_wrap(~language) +
    geom_point(size = 1, alpha = 0.5) +
    geom_smooth(method = "lm", formula = "y ~ x") +
    labs(
        y = "Levenshtein similarity", 
        x = "Lexical frequency (Zipf score)\nExtracted from SUBTLEX", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    theme(
        legend.position = "none",
        panel.grid.major.y = element_blank()
    )
```

## \# Phonemes by Levenshtein

```{r}
#| label: fig-items-nphon-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association betwen number of phonemes and phonological similarity"
items %>% 
    left_join(select(pool, item, language)) %>% 
    drop_na(freq) %>% 
    ggplot(aes(n_phon, lv, colour = language, fill = language)) +
    facet_wrap(~language) +
    geom_point(size = 1, alpha = 0.25) +
    geom_smooth(method = "lm", formula = "y ~ x") +
    labs(
        x = "# Phonemes", 
        y = "Levenshtein similarity", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    ) 
```

# Participants

```{r}
#| label: fig-participants-time
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Data at which responses were collected"
participants <- left_join(
    participants, 
    select(multilex_data$logs, id, time, edu_parent, dominant_language = dominance)
)

participants %>% 
    group_by(time_stamp, dominant_language) %>% 
    summarise(n = n(), .groups = "drop") %>% 
    group_by(dominant_language) %>% 
    mutate(n = cumsum(n)) %>% 
    ggplot(aes(time_stamp, n, colour = dominant_language, fill = dominant_language)) +
    geom_vline(xintercept = ymd("2020-03-15"), size = 0.5) +
    annotate(geom = "text", y = 350, x = ymd("2020-03-15"), label = "COVID-19 lockdown",
             angle = 90, vjust = 1.5, hjust = 1) +
    geom_line(size = 1) +
    labs(x = "Date", y = "Number of responses", colour = "Dominant language") +
    scale_y_continuous(breaks = seq(0, 400, 50), limits = c(0, 400)) +
    guides(colour = guide_legend(ncol = 2)) +
    theme(
        legend.position = "top",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
    )
```

## Age

@@

## Language profile

```{r}
#| label: fig-participants-lp
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Participant degree of exposure (DoE) to Catalan and Spanish"
participants %>% 
    mutate(
        age = round(age),
        id = paste(id, "(", time, ")"),
        rank = rank(doe_catalan)
    ) %>% 
    pivot_longer(
        starts_with("doe_"), 
        names_to = "language",
        values_to = "doe",
        names_prefix = "doe_",
        names_transform = list(language = str_to_sentence)
    ) %>% 
    ggplot() +
    aes(x = doe, y = reorder(id, -rank), fill = language) +
    geom_col(width = 1, position = position_fill(), colour = "white", size = 0.1) +
    geom_vline(xintercept = seq(0, 1, 0.25), colour = "grey") +
    labs(
        x = "Degree of Exposure (DoE)",
        y = "Participant",
        colour = "Language",
        fill = "Language"
    ) +
    scale_x_continuous(labels = percent) +
    theme(
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.title = element_blank()
    )
```

## SES/parental education

```{r}
#| label: tbl-participants-edu
#| tbl-cap: Participant exposure to the language (*DoE*) and socio-economic status (*SES*)
#| echo: true
#| message: false
#| warning: false
participants %>%
    mutate(
        edu_parent = fct_explicit_na(as.factor(edu_parent)),
        doe_catalan = cut(doe_catalan, breaks = seq(0, 1, 0.1), include.lowest = TRUE)
    ) %>% 
    count(doe_catalan, edu_parent) %>% 
    right_join(
        expand_grid(
            doe_catalan = levels(.$doe_catalan),
            edu_parent = unique(multilex_data$logs$edu_parent)
        )
    ) %>% 
    rename(total = n) %>% 
    pivot_wider(names_from = edu_parent, values_from = total, values_fill = 0) %>%
    mutate(total = rowSums(cbind(.[,3:8]), na.rm = TRUE)) %>% 
    rename(Total = total) %>% 
    select(-"NA") %>% 
    gt() %>% 
    cols_label(
        doe_catalan = "DoE Catalan"
    ) %>% 
    summary_rows(
        columns = 2:7, 
        fns = list(Total = ~sum(., na.rm = TRUE)),
        formatter = fmt_number,
        decimals = 0
    ) %>% 
    sub_missing(columns = 2:8, missing_text = "--") %>% 
    tab_spanner("Educational attainment", 3:8) %>% 
    tab_style(
        style = list(cell_borders(weight = px(2), sides = "left", color = "grey50")),
        locations = cells_body(columns = 9, rows = 1:10)
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(columns = 2:7)
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_labels(columns = c(1, 8))
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(spanners = "Educational attainment")
    )
```

```{r}
#| label: fig-participants-edu
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
participants %>%
    left_join(select(multilex_data$logs, id, time, edu_parent)) %>% 
    mutate(
        doe_2 = pmin(doe_catalan, doe_spanish),
        doe_2 = cut(doe_2, seq(0, 0.5, 0.1), include.lowest = TRUE)
    ) %>% 
    count(doe_2, edu_parent) %>% 
    ggplot() +
    aes(
        x = doe_2,
        y = n,
        fill = edu_parent
    ) +
    geom_col(position = position_fill()) +
    scale_fill_manual(values = sort(pal_d3()(6), decreasing = TRUE)) +
    labs(
        x = "Exposure to second language (L2 DoE)",
        y = "Proportion of participants",
        fill = "Highest parental educational attainment"
    ) +
    scale_y_continuous(labels = percent) +
    theme(
        legend.position = "none",
        panel.grid = element_blank()
    )
```

# Data analysis

## Model comparison

We initially fitted a base model (`model_fit_0`) that only included the predictors $Age$ (`age_std`), $Frequency$ (`freq_std`), $Phonemes$ (`n_phon_std`), and $DoE$ (`doe_std`) as nuisance parameters, along with random intercepts and slopes of all predictors by $Participant$ (`id`) and $Item$ (`te`), and all correlation parameters. We then expanded this model by adding the main effect of $Levenshtein$ (`lv_std`) (`model_fit_1`) and its random slope by $Participant$. Then, we added the $DoE \times Levenshtein$ (`doe_std:lv_std`) two-way interaction (`model_fit_2`) and its random slope by $Participant$. Finally, we added the $Age \times Levenshtein$ (`age_std:doe_std:lv_std`) three-way interaction with its random slope by $Participant$, along with a random slope for the $Age \times DoE$ (`age_std:doe_std`) two-way interaction.

## Model equation

@@

In addition, we specified a $LKJ$ prior--$\sim LKJ(2)$ for the correlations between group-level coefficients. We implemented the model prior following {brms} syntax as:

>>>>

```{r}
#| label: setup
#| include: false
tar_load_globals() # load packages and options
tar_load_everything() # load targets as R objects
theme_set(theme_custom()) # set custom ggplot theme

# resolve namespace conflicts
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
options(
    knitr.kable.NA = "-",
    knitr.duplicate.label = "allow",
    ggplot2.discrete.fill = ggsci::pal_d3()(4),
    ggplot2.discrete.colour = ggsci::pal_d3()(4),
    ggplot2.continuous.fill = ggplot2::scale_color_gradient,
    ggplot2.continuous.colour = ggplot2::scale_color_gradient
)
```

# Methods

Data and materials are available at OSF (https://osf.io/hy984/), and code is available at GitHub (https://github.com/gongcastro/trajectories).

## Questionnaires

```{r}
#| label: questionnaire
#| echo: true
#| message: false
#| warning: false
pool <- mutate(bvqdev::pool, language = ifelse(str_detect(item, "cat_"), "Catalan", "Spanish"))
n_categories <- distinct(pool, category) %>% nrow()
n_items <- length(unique(pool$item))
n_item_language <- count(pool, language) %>% 
    pull(n) %>%
    set_names(c("catalan", "spanish"))
```

The questionnaire was implemented on-line using the formR platform [@arslan2020formr], and was structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items and the Spanish inventory contained `r n_item_language["spanish"]` items. Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* \[dog\] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. When there were two acceptable translation equivalents for a given word, we included both in separate items (e.g., Catalan *acabar* \[*to finish*\] and Spanish *acabar* and *terminar*), or merged them into a single items (e.g., Spanish *mono* \[*monkey*\] and Catalan *mono/mico*. We included items from a diverse sample of `r n_categories` semantic/functional categories (see Appendix 1). For the analyses included in this study, we excluded items from the adverbs, auxiliary words, connectives, interjections and games and routines categories, so that only data from content words (nouns, adjectives, and verbs) were used.

@@

## Items

```{r}
#| label: tbl-items-summary
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Lexical frequencies. Mean, standard error, and 95% confidence interval of lexical frequencies of items included in the Catalan and Spanish lists, reported separately for identical cognates, non-identical cognates, and non-cognates."
items <- left_join(items, select(pool, te, item, language, class, category))

items %>%
    unnest(list) %>% 
    group_by(language, list) %>% 
    summarise_at(
        vars(freq_zipf, n_phon, lv), 
        list(
            n = ~sum(!is.na(.)),
            mean = ~mean(., na.rm = TRUE),
            sd = ~sd(., na.rm = TRUE),
            min = ~min(., na.rm = TRUE),
            max = ~max(., na.rm = TRUE)
        )
    ) %>% 
    select(-c(n_phon_n, lv_n)) %>% 
    rename(n = freq_zipf_n) %>% 
    gt(groupname_col = "language") %>% 
    fmt_number(matches("freq_|n_phon_mean|n_phon_sd")) %>% 
    fmt_percent(matches("lv_")) %>% 
    tab_spanner("Frequency (Zipf)", columns = starts_with("freq_")) %>% 
    tab_spanner("Length (Phonemes)", columns = starts_with("n_phon_")) %>% 
    tab_spanner("Phon. similarity (Levenshtein)", columns = starts_with("lv_")) %>% 
    cols_merge_uncert(col_val = freq_zipf_mean, col_uncert = freq_zipf_sd) %>% 
    cols_merge_uncert(col_val = n_phon_mean, col_uncert = n_phon_sd) %>% 
    cols_merge_uncert(col_val = lv_mean, col_uncert = lv_sd) %>% 
    cols_merge_range(col_begin = freq_zipf_min, col_end = freq_zipf_max) %>% 
    cols_merge_range(col_begin = n_phon_min, col_end = n_phon_max) %>% 
    cols_merge_range(col_begin = lv_min, col_end = lv_max) %>% 
    cols_label(
        list = "",
        n = md("*N*"),
        freq_zipf_mean = md("Mean (SD)"),
        freq_zipf_min = md("Range"),
        n_phon_mean = md("Mean (SD)"),
        n_phon_min = md("Range"),
        lv_mean = md("Mean (SD)"),
        lv_min = md("Range")
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(everything())
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(everything())
    ) %>% 
    summary_rows(
        fns = list("Grand mean" = ~mean(.)),
        columns = ends_with("_mean")
    ) %>% 
    summary_rows(
        fns = list("N" = ~sum(.)),
        columns = n,
        formatter = fmt_integer
    )
```

### Lexical frequency

```{r}
#| label: fig-items-freq
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Distribution of lexical frequency across items"
items %>% 
    ggplot() +
    aes(
        x = language,
        y = freq_zipf,
        fill = language,
        colour = language
    ) +
    stat_slab(slab_colour = "white", slab_size = 1, alpha = 0.5, scale = 0.5) +
    stat_dots(side = "left", scale = 0.5, position = position_nudge(x = -0.05)) + 
    stat_pointinterval(
        position = position_dodge(width = 0.1),
        point_size = 2.5,
        colour = "black"
    ) +
    labs(
        x = "Frequency (Zipf)",
        y = "Probability density",
        fill = "Language",
        shape = "Language",
        colour = "Language"
    ) +
    theme(
        legend.position = "none",
        legend.title = element_blank()
    )
```

### \# Phonemes

```{r}
#| label: fig-items-nphon
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Distribution of number of phonemes across items"
items %>%
    count(language, n_phon) %>% 
    ggplot() +
    aes(
        x = n_phon,
        y = n,
        fill = language,
        colour = language,
        shape = language
    ) +
    facet_wrap(~language, ncol = 1) +
    geom_col() +
    labs(
        x = "# Phonemes",
        y = "Number of words",
        fill = "Language",
        shape = "Language",
        colour = "Language"
    ) +
    scale_x_continuous(breaks = 1:15) +
    theme(
        legend.position = "none",
        axis.ticks.x = element_line(),
        legend.title = element_blank()
    )
```

### Levenshtein

```{r}
#| label: fig-items-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Distribution of Levenshtein similarities across items"
items %>% 
    distinct(te, .keep_all = TRUE) %>% 
    ggplot(aes(x = lv)) +
    geom_dots(fill = ggsci::pal_d3()(4)[1], colour = ggsci::pal_d3()(4)[1]) +
    labs(x = "Levenshtein distance", y = "Probability density") +
    scale_x_continuous(labels = percent) +
    theme(
        legend.position = "top"
    )
```

### Frequency by \# Phonemes

```{r}
#| label: fig-items-frequency-phonemes
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association betwen number of lexical frequency and number of phonemes"
items %>% 
    drop_na(freq_zipf) %>% 
    count(n_phon, freq_zipf, language) %>% 
    ggplot(aes(n_phon, freq_zipf, colour = language, fill = language)) +
    facet_wrap(~language) +
    geom_point(size = 2, alpha = 0.25) +
    geom_smooth(method = "lm", formula = "y ~ x") +
    labs(
        x = "# Phonemes", 
        y = "Lexical frequency (Zipf score)\nExtracted from SUBTLEX", 
        colour = "Language", 
        fill = "Language"
    ) +
    coord_equal(ratio = 10) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    ) 
```

### Frequency by Levenshtein

```{r}
#| label: fig-items-frequency-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association betwen lexical frequency and phonological similarity"
items %>% 
    left_join(select(pool, item, language)) %>% 
    drop_na(freq_zipf) %>% 
    count(freq_zipf, lv, language) %>% 
    ggplot(aes(freq_zipf, lv, colour = language, fill = language, size = n)) +
    facet_wrap(~language) +
    geom_point(size = 2, alpha = 0.25) +
    geom_smooth(method = "lm", formula = "y ~ x") +
    labs(
        y = "Levenshtein similarity", 
        x = "Lexical frequency (Zipf score)\nExtracted from SUBTLEX", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    theme(
        legend.position = "none",
        panel.grid.major.y = element_blank()
    )
```

### \# Phonemes by Levenshtein

```{r}
#| label: fig-items-nphon-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association betwen number of phonemes and phonological similarity"
items %>% 
    left_join(select(pool, item, language)) %>% 
    drop_na(freq_zipf) %>% 
    count(n_phon, lv, language) %>% 
    ggplot(aes(n_phon, lv, colour = language, fill = language, size = n)) +
    facet_wrap(~language) +
    geom_point(shape = 1, stroke = 1) +
    geom_smooth(method = "lm", formula = "y ~ x", size = 1) +
    labs(
        x = "# Phonemes", 
        y = "Levenshtein similarity", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    ) 
```

## Participants

```{r}
#| label: fig-participants-time
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Data at which responses were collected"
participants %>% 
    group_by(time_stamp) %>% 
    summarise(n = n(), .groups = "drop") %>% 
    mutate(n = cumsum(n)) %>% 
    ggplot(aes(time_stamp, n)) +
    geom_vline(xintercept = ymd("2020-03-15"), size = 0.5) +
    annotate(geom = "text", y = 350, x = ymd("2020-03-15"), label = "COVID-19 lockdown",
             angle = 90, vjust = 1.5, hjust = 1) +
    geom_line(size = 2, colour = pal_d3()(1)[1]) +
    labs(x = "Date", y = "Number of responses", colour = "Dominant language") +
    scale_y_continuous(breaks = seq(0, 400, 50), limits = c(0, 400)) +
    scale_x_date(
        date_breaks = "3 month", 
        date_labels = "%B %Y"
    ) +
    guides(colour = guide_legend(ncol = 2)) +
    theme(
        legend.position = "top",
        legend.title = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 20, hjust = 1),
        axis.title.x = element_blank(),
    )
```

### Age

@@

### Language profile

```{r}
#| label: fig-participants-lp
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Participant degree of exposure (DoE) to Catalan and Spanish"
participants %>% 
    mutate(
        age = round(age),
        id = paste(id, "(", time, ")"),
        rank = rank(doe_catalan)
    ) %>% 
    pivot_longer(
        starts_with("doe_"), 
        names_to = "language",
        values_to = "doe",
        names_prefix = "doe_",
        names_transform = list(language = str_to_sentence)
    ) %>% 
    ggplot() +
    aes(x = doe, y = reorder(id, -rank), fill = language) +
    geom_col(width = 1, position = position_fill(), colour = "white", size = 0.1) +
    geom_vline(xintercept = seq(0, 1, 0.25), colour = "grey") +
    labs(
        x = "Degree of Exposure (DoE)",
        y = "Participant",
        colour = "Language",
        fill = "Language"
    ) +
    scale_x_continuous(labels = percent) +
    theme(
        legend.position = "top",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank()
    )
```

### SES/parental education

```{r}
#| label: tbl-participants-edu
#| tbl-cap: Participant exposure to the language (*DoE*) and socio-economic status (*SES*)
#| echo: true
#| message: false
#| warning: false
participants %>%
    left_join(bvq_data$logs) %>% 
    mutate(
        edu_parent = fct_explicit_na(as.factor(edu_parent)),
        doe_catalan = cut(doe_catalan, breaks = seq(0, 1, 0.2), include.lowest = TRUE)
    ) %>% 
    count(doe_catalan, edu_parent) %>% 
    right_join(
        expand_grid(
            doe_catalan = levels(.$doe_catalan),
            edu_parent = unique(bvq_data$logs$edu_parent)
        )
    ) %>% 
    rename(total = n) %>% 
    pivot_wider(names_from = edu_parent, values_from = total, values_fill = 0) %>%
    mutate(total = rowSums(cbind(.[,3:8]), na.rm = TRUE)) %>% 
    rename(Total = total) %>% 
    select(-"NA") %>% 
    gt() %>% 
    summary_rows(
        columns = Complementary:Primary, 
        fns = list(Total = ~sum(., na.rm = TRUE)),
        formatter = fmt_integer
    ) %>% 
    cols_label(doe_catalan = "DoE Catalan") %>%
    sub_missing(columns = 2:8, missing_text = "--") %>% 
    tab_spanner("Educational attainment", Complementary:Primary) %>% 
    tab_style(
        style = list(cell_borders(weight = px(2), sides = "left", color = "grey50")),
        locations = cells_body(columns = 9, rows = 1:5)
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(columns = 2:7)
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_labels(columns = c(1, 8))
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(spanners = "Educational attainment")
    )
```

## Exposure-weighted lexical frequency

We created a predictor ($Exposure$) to account for the exposure rate of each child to each of the words their parents responded to, weighted by the child's exposure to the language the word belongs to. This measure is the product of the lexical frequency of the word (Zipf score) and the child's degree of exposure to the corresponding language (a proportion):

$$
\textbf{Exposure}_{ij} = \textbf{Frequency}_i \times \textbf{DoE}_{j}
$$

For instance, for a child who is reportedly exposure to Catalan 80% of the time, and to Spanish 20% of the time, the expected exposure to the word *cavall* (horse, in Catalan, with a lexical frequency of `r round(items$freq_zipf[items$item=="cat_cavall"], 2)`) would be `r round(items$freq_zipf[items$item=="cat_cavall"]*0.80, 2)`, while that of its translation to Spanish *caballo* would be `r round(items$freq_zipf[items$item=="cat_cavall"]*0.20, 2)`:

$$
\begin{aligned}
\textbf{Exposure}_{\textbf{cavall, Catalan}} &= \textbf{Frequency}_{cavall} \times \textbf{DoE}_{Catalan} = 6.14 \times 0.8 = 4.91 \\
\textbf{Exposure}_{\textbf{caballo, Spanish}} &= \textbf{Frequency}_{caballo} \times \textbf{DoE}_{Spanish} = 6.14 \times 0.2 = 1.23
\end{aligned}
$$

```{r}
#| label: fig-exposure-distribution
df %>% 
    mutate(language = ifelse(grepl("cat_", item), "Catalan", "Spanish")) %>% 
    ggplot(aes(exposure, fill = language)) +
    facet_wrap(~language) +
    geom_histogram(bins = 10, colour = "white", size = 1, show.legend = FALSE) +
    labs(x = "Exposure (Frequency \u00d7 DoE)", y = "Number of observations") +

    df %>% 
    mutate(language = ifelse(grepl("cat_", item), "Catalan", "Spanish")) %>% 
    ggplot(aes(x = freq, y = exposure, colour = language)) +
    facet_wrap(~language) + 
    geom_point(size = 0.1, alpha = 0.5, show.legend = FALSE) +
    labs(x = "Frequency (Zipf score)", y = "Exposure") +
    plot_layout(ncol = 1)
```

## Data analysis

### Model comparison

We initially fitted a base model (`model_fit_0`) that only included the predictors $Age$ (`age_std`), $Phonemes$ (`n_phon_std`), along with random intercepts and slopes of all predictors by $Participant$ (`id`) and $Item$ (`te`), and all correlation parameters. We then expanded this model by adding the main effect of $Exposure$ (`exposure`) (`model_fit_1`), and then $Levenshtein$ (`lv_std`) (`model_fit_2`) and their corresponding random slopes by $Participant$. Then, we added the $Exposure \times Levenshtein$ (`exposure_std:lv_std`) two-way interaction (`model_fit_3`) and its random slope by $Participant$. Finally, we added the $Age \times Exposure \times Levenshtein$ (`age_std:exposure_std:lv_std`) three-way interaction with its random slope by $Participant$, along with a random slope for the $Age \times Exposure$ (`age_std:exposure_std`) two-way interaction.

### Model equation

@@

In addition, we specified a $LKJ$ prior--$\sim LKJ(2)$ for the correlations between group-level coefficients.

#### Prior predictions

::: {.panel-tabset}

##### Prior-predicted mean

```{r}
#| label: fig-prior-mean
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Expected prior predictions"
nd <- expand.grid(
    n_phon_std = 0,
    age_std = (10:36-mean(df$age))/(sd(df$age)),
    exposure_std = 0,
    lv_std = 0
)

m <- add_epred_draws(nd, model_fit_4_prior, ndraws = NULL, re_formula = NA) %>% 
    filter(.category != "No") %>% 
    pivot_wider(names_from = ".category", values_from = ".epred") %>% 
    mutate(Understands = Understands + `Understands and Says`) %>% 
    pivot_longer(
        c(Understands, `Understands and Says`),
        names_to = ".category",
        values_to = ".epred"
    ) %>% 
    mutate(
        .category = factor(
            .category,
            levels = c("Understands and Says", "Understands"),
            labels = c("Production\n(Understands and Says))", "Comprehension\n(Understands)")
        )
    )
